---
title: "MAS61004 - Assessed Project"
author: "Allan Cousins, Matthew Knowles, Oliver Hewitt"
date: "2/23/2023"
output:
  pdf_document: 
    number_sections: true
  html_document: 
    number_sections: true
fontsize: 11pt
urlcolor: blue
header-includes:
  - \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.align = 'center')
library(tidyverse)
library(ggplot2)
accidents <- read.csv("dft-road-casualty-statistics-accident-2020.csv")
vehicles <- read.csv("dft-road-casualty-statistics-casualty-2020.csv")
casualties <- read.csv("dft-road-casualty-statistics-vehicle-2020.csv")
```

# Executive Summary

The main purpose of this investigation is to describe the relationship primarily between accident severity and speed limit, and other factors included in the data provided by the UK Government's Department for Transport.

Our focus is to assess the risk of a fatal or severe injury, given that one has already occurred - and so the data in concern is road safety data exclusively from the year 2020. Analysis in the report will be carried out in R using Generalised Linear Models (GLMs), as it would be incorrect to assume the data is normally distributed.

(Outline conclusions here)

# Method

Our primary analysis, as outlined in the summary, will be GLMs fitted to the data in R. The data provided are public domain, and include: "Road Safety Data - Accidents 2020.csv", "Road Safety Data - Vehicles 2020.csv" and "Road Safety Data - Casualties 2020.csv", merged by an accident index present in all three data sets.

As mentioned previously, we wouldn't assume the data are normally distributed nor would we expect the outcome of our residuals to be normally distributed. We assume that a large demographic of the UK were not involved in a traffic accident in 2020, and an even smaller demographic were involved once or multiple times. In addition, not all response variables considered in the analysis are continuous - hence GLM would be an appropriate method to use under this non-normality and non-continuous response assumption.

With this established, we determined which link function to use in the GLM using the ggplot package (see Appendix, A1) to explore the data. There doesn't appear to be an immediate pattern or linearity in terms of the effect of speed limit on accident count, and so we may choose a standard Poisson Regression model, which is best used for modelling count data. Additionally, the Poisson Distribution is most commonly used to find the probability of events occurring within a given time interval. [1]  We use the canonical link function, which in the case of Poisson regression is the log-link function.


(Talk about the 3 components of GLM here)

\newpage
# Results

## Descriptive plots:

```{r plot, echo = FALSE, fig.cap="Bar chart presenting Number of Accident by Day"}
week_count <- accidents %>%
    group_by(day_of_week, accident_severity) %>%
    count()
week_count$day_of_week <- as.factor(week_count$day_of_week)
week_count$accident_severity <- as.factor(week_count$accident_severity)
levels(week_count$day_of_week) <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
ggplot(week_count, aes(x = day_of_week, y = n, fill = accident_severity)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("#999999", "#E69F00", "#56B4E9")) + xlab("Day of the Week") + ylab("Number of Accidents")
```
## GLM:

We use the standard function _summary()_ to produce result summaries of our model which uses the _glm()_ function (see Appendix, A2 and A3).

A3 is our summary of the _glm()_ function. We use this output to see which predictor variables have an effect on the response variable by the _p_ value seen in the final column _Pr(>|z|)_ - the predictor variable has an effect on the response when _p < 0.05_.

Denote the significance of these variables by the number of stars on each row: we see that the number of vehicles involved in the accident, road type and speed limit bear the most significance on the outcome. Following not too far behind is junction detail and light conditions. Then, bearing even less significance is weather conditions. The variables with little to no effect on the response variables are: road surface conditions, first road class, second road class, and the day of the week.

Ultimately, of all the predictors compared within the analysis, speed limit has the smallest _p_ value which implies a stronger relationship between accident severity and speed limit overall.


\newpage
# Appendix

A1: Log Accident Count by Speed Limit

```{r, echo = FALSE}
speed_limit_count <- accidents %>%
    group_by(speed_limit, accident_severity) %>%
    count() %>%
    mutate(logn = log(n))
speed_limit_count$speed_limit <- as.factor(speed_limit_count$speed_limit)
speed_limit_count$accident_severity <- as.factor(speed_limit_count$accident_severity)
ggplot(speed_limit_count %>% filter(speed_limit != -1), aes(shape = accident_severity, colour = accident_severity)) +
    geom_point(aes(x = speed_limit, y = logn), size = 3) +
    xlab("Speed Limit") + ylab("Log Accident Count") 
```

A2: _glm_ function in R using the _poisson_ argument
```{r GLM code, echo = TRUE}
# Indices of variables we want to use
use_in_formula <- c(
    9, 10, 13, 18, 20, 21, 22, 24, 28, 29, 30)
# Subset the data by key variables
accidents_key <- accidents[, use_in_formula]
# Which observations have negatives in any column?
has.neg <- apply(accidents_key, 1, function (row) any(row < 0))
# Remove observations with negative values
accidents_data <- accidents_key[-which(has.neg), ]
# Fit GLM
model <- glm(data = accidents_data,
             formula = accident_severity ~ .,
             family = poisson(link = "log"))
```

\newpage
A3: Summary of the _poisson_ model
```{r, echo = FALSE}
summary(model)
```



# References

[1] "Tutorial: Poisson Regression in R" - dataquest, created February 27th 2019.
https://www.dataquest.io/blog/tutorial-poisson-regression-in-r/
